---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
  pdf_document: default
---
# First of all we need to load and preprocess the given data an then answer the questions.

```{r load}
dataActivity <- read.csv(file="activity.csv")
dataActivity$date <- as.Date(dataActivity$date)
```

## What is mean total number of steps taken per day?

1.- Calculate the total number of steps taken by day

```{r steps_day}

library("dplyr")
library("ggplot2")

dayStepCount <- dataActivity %>% group_by(date) %>% summarise(day.count = sum(steps, na.rm=TRUE))
```

2.- Make a histogram of the total number of steps taken each day, then calculate and report the mean and median of the total number of steps taken per day.

```{r mean_median}
mean(dayStepCount$day.count, na.rm=T)
median(dayStepCount$day.count, na.rm=T)

ggplot(dayStepCount) + geom_histogram(aes(x = day.count), binwidth = 1000) +
        theme_bw() + ggtitle('Total number of steps per day') + 
        xlab('Number of steps per day')
```

## What is the average daily activity pattern?

1.- Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
tidyData <- dataActivity[!is.na(dataActivity$date) & !is.na(dataActivity$steps),]
plotData <- tapply(tidyData$steps, tidyData$interval, mean, na.rm=T)
plot(names(plotData), plotData,type="l", xlab = "Interval", ylab = "Average Steps")
```

2.-Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
maxInterval<-which.max(plotData)
maxInterval
```

## Imputing missing values

1.- Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r NAs}
summary(dataActivity)
```

2.- With this simple function we know that there are 2304 NA's only in the steps field, and now we can proceed to impute missing values, the strategy that I'm going to use is use the mean of steps taken at each interval.

```{r}
# Mean step count per interval
MeanActivity <- dataActivity %>% group_by(interval) %>% 
summarise(meanSteps = mean(steps, na.rm = T))
dataActivity.Imputed <- left_join(dataActivity, MeanActivity)
# If step is NA change to meansteps
dataActivity.Imputed <- dataActivity.Imputed %>% mutate(steps = ifelse(is.na(steps),round(meanSteps,0),steps))
```

3.- The new DataSet with imputed data is dataActivity.Imputed shown in the past chunk of code.

4.- Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r histogram}
# Number of steps/day imputed
dayStepCount.imp <- dataActivity.Imputed %>% group_by(date) %>% 
summarise(day.count = sum(steps))
# Mean and median number of steps/day imputed
meanStepCount.imp <- dayStepCount.imp %>% ungroup() %>% 
summarise(mean = mean(day.count, na.rm = T), median = median(day.count, na.rm = T))

ggplot(dayStepCount.imp) + geom_histogram(aes(x = day.count), binwidth = 1000) + 
        ggtitle('Histogram of total number of steps taken each day\nAfter imputation') + 
        theme_bw() + xlab('Daily step count')
```

The comparison between means and median of the imputed and original data sets are almost the same, the strategy is working as expected.

```{r}
## Imputed DS
meanStepCount.imp %>% data.frame
## Original DS
mean(dayStepCount$day.count, na.rm=T)
median(dayStepCount$day.count, na.rm=T)
```

## Are there differences in activity patterns between weekdays and weekends?

1.- Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r weekdays/weekend}
# set the locale for english weekdays
Sys.setlocale("LC_TIME", "English")
#Make dates dateobjects and generate weekday field
dataActivity.wd <- dataActivity.Imputed %>% mutate(date = as.Date(date), weekDay = weekdays(date)) %>%
# Calculate if weekend or weekday to create factor
mutate(weekEnd.day = ifelse(weekDay %in% c('Saturday','Sunday'),'weekEnd','weekDay'))
# Reorder weekend/weekday factor
dataActivity.wd$weekEnd.day <- relevel(factor(dataActivity.wd$weekEnd.day), ref = 'weekEnd')
# Average number of steps during weekend/weekdays
dataActivity.wd.mean <- dataActivity.wd %>% group_by(interval,weekEnd.day) %>%
summarise(mean.steps = mean(steps, na.rm = T))
```

2.- Make a panel plot containing a time series plot (i.e.  type = "l" ) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). The plot should look something like the following, which was created using simulated data:

```{r Plot WE & Wd}
ggplot(dataActivity.wd.mean) + geom_line(aes(x = interval, y = mean.steps)) + 
        facet_wrap(~weekEnd.day,nrow = 2) + theme_bw() + ggtitle('WeekEnd & WeekDay Patterns') + 
        xlab('Interval') + ylab('Mean number of steps')
```

We can notate differences in the patterns of the weekdays and weekend, and certain similarities at the beggining of the interval.